---
title: '1. Neighborhood Deprivation Indices'
author: 'Ian D. Buller (GitHub: @idblr)'
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Neighborhood Deprivation Indices}
  %\VignetteEngine{R.rsp::asis}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = FALSE, fig.show = 'hold')
```

Start with the necessary packages for the vignette.

```{r packages, results = 'hide'}
loadedPackages <- c('dplyr', 'ggplot2', 'ndi', 'sf', 'tidycensus', 'tigris')
invisible(lapply(loadedPackages, library, character.only = TRUE))
options(tigris_use_cache = TRUE)
```

Set your U.S. Census Bureau access key. Follow [this link](http://api.census.gov/data/key_signup.html) to obtain one. Specify your access key in the `messer()` or `powell_wiley()` functions using the `key` argument of the `get_acs()` function from the [*tidycensus*](https://CRAN.R-project.org/package=tidycensus) package called within each or by using the `census_api_key()` function from the [*tidycensus*](https://CRAN.R-project.org/package=tidycensus) package before running the `messer()` or `powell_wiley()` functions (see an example of the latter below).

```{r access_key_private, echo = FALSE}
source(file.path('..', 'dev', 'private_key.R'))
census_api_key(private_key)
```

```{r access_key_public, eval = FALSE}
census_api_key('...') # INSERT YOUR OWN KEY FROM U.S. CENSUS API
```

## Compute *NDI* (Messer)

Compute the *NDI* (Messer) values (2006-2010 5-year ACS) for Georgia, U.S.A., census tracts. This metric is based on [Messer et al. (2006)](https://doi.org/10.1007/s11524-006-9094-x) with the following socio-economic status (SES) variables: 

| Characteristic | SES dimension | ACS table source | Description |
| -------------- | ------------- | ---------------- | ----------- |
| OCC | Occupation | C24030 | Percent males in management, science, and arts occupation |
| CWD | Housing | B25014 | Percent of crowded housing |
| POV | Poverty | B17017 | Percent of households in poverty |
| FHH | Poverty | B25115 | Percent of female headed households with dependents |
| PUB | Poverty | B19058 | Percent of households on public assistance |
| U30 | Poverty | B19001 | Percent households earning <$30,000 per year |
| EDU | Education | B06009 | Percent earning less than a high school education |
| EMP | Employment | B23001 (2010 only); B23025 (2011 onward) | Percent unemployed |

```{r messer, results = 'hide'}
messer2010GA <- messer(state = 'GA', year = 2010, round_output = TRUE)
```

One output from the `messer()` function is a tibble containing the identification, geographic name, *NDI* (Messer) values, and raw census characteristics for each tract.

```{r messer_out1}
messer2010GA$ndi
```

A second output from the `messer()` function is the results from the principal component analysis used to compute the *NDI* (Messer) values.

```{r messer_out2}
messer2010GA$pca
```

A third output from the `messer()` function is a tibble containing a breakdown of the missingness of the census characteristics used to compute the *NDI* (Messer) values.

```{r messer_out3}
messer2010GA$missing
```

We can visualize the *NDI* (Messer) values geographically by linking them to spatial information from the [*tigris*]([*tidycensus*](https://CRAN.R-project.org/package=tigris) package and plotting with the [*ggplot2*]([*tidycensus*](https://CRAN.R-project.org/package=ggplot2) package suite.

```{r messer_prep, results = 'hide'}
# Obtain the 2010 counties from the 'tigris' package
county2010GA <- counties(state = 'GA', year = 2010, cb = TRUE)
# Remove first 9 characters from GEOID for compatibility with tigris information
county2010GA$GEOID <- substring(county2010GA$GEO_ID, 10) 

# Obtain the 2010 census tracts from the 'tigris' package
tract2010GA <- tracts(state = 'GA', year = 2010, cb = TRUE)
# Remove first 9 characters from GEOID for compatibility with tigris information
tract2010GA$GEOID <- substring(tract2010GA$GEO_ID, 10) 

# Join the NDI (Messer) values to the census tract geometry
GA2010messer <- tract2010GA %>%
  left_join(messer2010GA$ndi, by = 'GEOID')
```

```{r messer_plot, fig.height = 7, fig.width = 7}
# Visualize the NDI (Messer) values (2006-2010 5-year ACS) for Georgia, U.S.A., census tracts 
## Continuous Index
ggplot() +
  geom_sf(
    data = GA2010messer,
    aes(fill = NDI),
    size = 0.05,
    color = 'transparent'
  ) +
  geom_sf(
    data = county2010GA,
    fill = 'transparent',
    color = 'white',
    size = 0.2
  ) +
  theme_minimal() +
  scale_fill_viridis_c() +
  labs(fill = 'Index (Continuous)', caption = 'Source: U.S. Census ACS 2006-2010 estimates') +
  ggtitle(
    'Neighborhood Deprivation Index (Messer)',
    subtitle = 'GA census tracts as the referent'
  )

## Categorical Index
### Rename '9-NDI not avail' level as NA for plotting
GA2010messer$NDIQuartNA <-
  factor(
    replace(
      as.character(GA2010messer$NDIQuart),
      GA2010messer$NDIQuart == '9-NDI not avail',
      NA
    ),
    c(levels(GA2010messer$NDIQuart)[-5], NA)
  )

ggplot() +
  geom_sf(
    data = GA2010messer,
    aes(fill = NDIQuartNA),
    size = 0.05,
    color = 'transparent'
  ) +
  geom_sf(
    data = county2010GA,
    fill = 'transparent',
    color = 'white',
    size = 0.2
  ) +
  theme_minimal() +
  scale_fill_viridis_d(guide = guide_legend(reverse = TRUE), na.value = 'grey80') +
  labs(fill = 'Index (Categorical)', caption = 'Source: U.S. Census ACS 2006-2010 estimates') +
  ggtitle(
    'Neighborhood Deprivation Index (Messer) Quartiles',
    subtitle = 'GA census tracts as the referent'
  )
```

The results above are at the tract level. The *NDI* (Messer) values can also be calculated at the county level. 

```{r messer_county_prep, results = 'hide'}
messer2010GA_county <- messer(geo = 'county', state = 'GA', year = 2010)

# Join the NDI (Messer) values to the county geometry
GA2010messer_county <- county2010GA %>%
  left_join(messer2010GA_county$ndi, by = 'GEOID')
```

```{r messer_county_plot, fig.height = 7, fig.width = 7}
# Visualize the NDI (Messer) values (2006-2010 5-year ACS) for Georgia, U.S.A., counties
## Continuous Index
ggplot() +
  geom_sf(
    data = GA2010messer_county,
    aes(fill = NDI),
    size = 0.20,
    color = 'white'
  ) +
  theme_minimal() +
  scale_fill_viridis_c() +
  labs(fill = 'Index (Continuous)', caption = 'Source: U.S. Census ACS 2006-2010 estimates') +
  ggtitle(
    'Neighborhood Deprivation Index (Messer)',
    subtitle = 'GA counties as the referent'
  )

## Categorical Index
### Rename '9-NDI not avail' level as NA for plotting
GA2010messer_county$NDIQuartNA <-
  factor(
    replace(
      as.character(GA2010messer_county$NDIQuart),
      GA2010messer_county$NDIQuart == '9-NDI not avail',
      NA
    ),
    c(levels(GA2010messer_county$NDIQuart)[-5], NA)
  )

ggplot() +
  geom_sf(
    data = GA2010messer_county,
    aes(fill = NDIQuartNA),
    size = 0.20,
    color = 'white'
  ) +
  theme_minimal() +
  scale_fill_viridis_d(guide = guide_legend(reverse = TRUE), na.value = 'grey80') +
  labs(fill = 'Index (Categorical)', caption = 'Source: U.S. Census ACS 2006-2010 estimates') +
  ggtitle(
    'Neighborhood Deprivation Index (Messer) Quartiles',
    subtitle = 'GA counties as the referent'
  )
```

## Compute *NDI* (Powell-Wiley)

Compute the *NDI* (Powell-Wiley) values (2016-2020 5-year ACS) for Maryland, Virginia, Washington, D.C., and West Virginia, U.S.A., census tracts. This metric is based on [Andrews et al. (2020)](https://doi.org/10.1080/17445647.2020.1750066) and [Slotman et al. (2022)](https://doi.org/10.1016/j.dib.2022.108002) with socio-economic status (SES) variables chosen by [Roux and Mair (2010)](https://doi.org/10.1111/j.1749-6632.2009.05333.x):

| Characteristic | SES dimension | ACS table source | Description |
| -------------- | ------------- | ---------------- | ----------- |
| MedHHInc | Wealth and income | B19013 | Median household income (dollars) |
| PctRecvIDR | Wealth and income | B19054 | Percent of households receiving dividends, interest, or rental income |
| PctPubAsst | Wealth and income | B19058 | Percent of households receiving public assistance |
| MedHomeVal | Wealth and income | B25077 | Median home value (dollars) |
| PctMgmtBusSciArt | Occupation | C24060 | Percent in a management, business, science, or arts occupation |
| PctFemHeadKids | Occupation | B11005 | Percent of households that are female headed with any children under 18 years |
| PctOwnerOcc | Housing conditions | DP04 | Percent of housing units that are owner occupied |
| PctNoPhone | Housing conditions | DP04 | Percent of households without a telephone |
| PctNComPlmb | Housing conditions | DP04 | Percent of households without complete plumbing facilities |
| PctEducHSPlus | Education | S1501 | Percent with a high school degree or higher (population 25 years and over) |
| PctEducBchPlus | Education | S1501 | Percent with a college degree or higher (population 25 years and over) |
| PctFamBelowPov | Wealth and income | S1702 | Percent of families with incomes below the poverty level |
| PctUnempl | Occupation | S2301 | Percent unemployed |

More information about the [codebook](https://gis.cancer.gov/research/NeighDeprvIndex_Methods.pdf) and [computation](https://gis.cancer.gov/research/NeighDeprvIndex_Methods.pdf) of the *NDI* (Powell-Wiley) can be found on a [GIS Portal for Cancer Research](https://gis.cancer.gov/research/files.html#soc-dep) website.

```{r powell_wiley, results = 'hide'}
powell_wiley2020DMVW <- powell_wiley(
  state = c('DC', 'MD', 'VA', 'WV'),
  year = 2020,
  round_output = TRUE
)
```

One output from the `powell_wiley()` function is a tibble containing the identification, geographic name, *NDI* (Powell-Wiley) values, and raw census characteristics for each tract.

```{r powell_wiley_out1}
powell_wiley2020DMVW$ndi
```

A second output from the `powell_wiley()` function is the results from the principal component analysis used to compute the *NDI* (Powell-Wiley) values.

```{r powell_wiley_out2}
powell_wiley2020DMVW$pca
```

A third output from the `powell_wiley()` function is a tibble containing a breakdown of the missingness of the census characteristics used to compute the *NDI* (Powell-Wiley) values.

```{r powell_wiley_out3}
powell_wiley2020DMVW$missing
```

A fourth output from the `powell_wiley()` function is a character string or numeric value of a standardized Cronbach's alpha. A value greater than 0.7 is desired.

```{r powell_wiley_out4}
powell_wiley2020DMVW$cronbach
```

We can visualize the *NDI* (Powell-Wiley) values geographically by linking them to spatial information from the [*tigris*]([*tidycensus*](https://CRAN.R-project.org/package=tigris) package and plotting with the [*ggplot2*]([*tidycensus*](https://CRAN.R-project.org/package=ggplot2) package suite.

```{r powell_wiley_prep, results = 'hide'}
# Obtain the 2020 counties from the 'tigris' package
county2020 <- counties(cb = TRUE)
county2020DMVW <- county2020[county2020$STUSPS %in% c('DC', 'MD', 'VA', 'WV'), ]

# Obtain the 2020 census tracts from the 'tigris' package
tract2020D <- tracts(state = 'DC', year = 2020, cb = TRUE)
tract2020M <- tracts(state = 'MD', year = 2020, cb = TRUE)
tract2020V <- tracts(state = 'VA', year = 2020, cb = TRUE)
tract2020W <- tracts(state = 'WV', year = 2020, cb = TRUE)
tracts2020DMVW <- rbind(tract2020D, tract2020M, tract2020V, tract2020W)

# Join the NDI (Powell-Wiley) values to the census tract geometry
DMVW2020pw <- tracts2020DMVW %>%
  left_join(powell_wiley2020DMVW$ndi, by = 'GEOID')
```

```{r powell_wiley_plot, fig.height = 4, fig.width = 7}
# Visualize the NDI (Powell-Wiley) values (2016-2020 5-year ACS) 
## Maryland, Virginia, Washington, D.C., and West Virginia, U.S.A., census tracts 
## Continuous Index
ggplot() +
  geom_sf(
    data = DMVW2020pw,
    aes(fill = NDI),
    color = NA
  ) +
  geom_sf(
    data = county2020DMVW,
    fill = 'transparent',
    color = 'white'
  ) +
  theme_minimal() +
  scale_fill_viridis_c(na.value = 'grey80') +
  labs(fill = 'Index (Continuous)', caption = 'Source: U.S. Census ACS 2016-2020 estimates') +
  ggtitle(
    'Neighborhood Deprivation Index (Powell-Wiley)',
    subtitle = 'DC, MD, VA, and WV tracts as the referent'
  )

## Categorical Index (Population-weighted quintiles)
### Rename '9-NDI not avail' level as NA for plotting
DMVW2020pw$NDIQuintNA <-
  factor(replace(
    as.character(DMVW2020pw$NDIQuint),
    DMVW2020pw$NDIQuint == '9-NDI not avail',
    NA
  ),
  c(levels(DMVW2020pw$NDIQuint)[-6], NA))

ggplot() +
  geom_sf(data = DMVW2020pw, aes(fill = NDIQuintNA), color = NA) +
  geom_sf(data = county2020DMVW, fill = 'transparent', color = 'white') +
  theme_minimal() +
  scale_fill_viridis_d(guide = guide_legend(reverse = TRUE), na.value = 'grey80') +
  labs(fill = 'Index (Categorical)', caption = 'Source: U.S. Census ACS 2016-2020 estimates') +
  ggtitle(
    'Neighborhood Deprivation Index (Powell-Wiley) Population-weighted Quintiles',
    subtitle = 'DC, MD, VA, and WV tracts as the referent'
  )
```

Like the *NDI* (Messer), we also compute county-level *NDI* (Powell-Wiley).

```{r powell_wiley_county_prep, results = 'hide'}
# Obtain the 2020 counties from the 'tigris' package
county2020DMVW <- counties(state = c('DC', 'MD', 'VA', 'WV'), year = 2020, cb = TRUE)

# NDI (Powell-Wiley) at the county level (2016-2020)
powell_wiley2020DMVW_county <- powell_wiley(
  geo = 'county',
  state = c('DC', 'MD', 'VA', 'WV'),
  year = 2020
)

# Join the NDI (Powell-Wiley) values to the county geometry
DMVW2020pw_county <- county2020DMVW %>%
  left_join(powell_wiley2020DMVW_county$ndi, by = 'GEOID')
```

```{r powell_wiley_county_plot, fig.height = 4, fig.width = 7}
# Visualize the NDI (Powell-Wiley) values (2016-2020 5-year ACS)
## Maryland, Virginia, Washington, D.C., and West Virginia, U.S.A., counties
## Continuous Index
ggplot() +
  geom_sf(
    data = DMVW2020pw_county,
    aes(fill = NDI),
    size = 0.20,
    color = 'white'
  ) +
  theme_minimal() +
  scale_fill_viridis_c() +
  labs(fill = 'Index (Continuous)', caption = 'Source: U.S. Census ACS 2016-2020 estimates') +
  ggtitle(
    'Neighborhood Deprivation Index (Powell-Wiley)',
    subtitle = 'DC, MD, VA, and WV counties as the referent'
  )

## Categorical Index
### Rename '9-NDI not avail' level as NA for plotting
DMVW2020pw_county$NDIQuintNA <-
  factor(
    replace(
      as.character(DMVW2020pw_county$NDIQuint),
      DMVW2020pw_county$NDIQuint == '9-NDI not avail',
      NA
    ),
    c(levels(DMVW2020pw_county$NDIQuint)[-6], NA)
  )

ggplot() +
  geom_sf(
    data = DMVW2020pw_county,
    aes(fill = NDIQuint),
    size = 0.20,
    color = 'white'
  ) +
  theme_minimal() +
  scale_fill_viridis_d(guide = guide_legend(reverse = TRUE), na.value = 'grey80') +
  labs(fill = 'Index (Categorical)', caption = 'Source: U.S. Census ACS 2016-2020 estimates') +
  ggtitle(
    'Neighborhood Deprivation Index (Powell-Wiley) Population-weighted Quintiles',
    subtitle = 'DC, MD, VA, and WV counties as the referent'
  )
```

### Advanced Features

#### Imputing missing census variables

In the `messer()` and `powell_wiley()` functions, missing census characteristics can be imputed using the `missing` and `impute` arguments of the `pca()` function in the [*psych*](https://CRAN.R-project.org/package=psych) package called within the `messer()` and `powell_wiley()` functions. Impute values using the logical `imp` argument (currently only calls `impute = 'median'` by default, which assigns the median values of each missing census variable for a geography).

```{r powell_wiley_imp, results = 'hide'}
powell_wiley2020DC <- powell_wiley(state = 'DC', year = 2020) # without imputation
powell_wiley2020DCi <- powell_wiley(state = 'DC', year = 2020, imp = TRUE) # with imputation

table(is.na(powell_wiley2020DC$ndi$NDI)) # n=13 tracts without NDI (Powell-Wiley) values
table(is.na(powell_wiley2020DCi$ndi$NDI)) # n=0 tracts without NDI (Powell-Wiley) values

# Obtain the 2020 census tracts from the 'tigris' package
tract2020DC <- tracts(state = 'DC', year = 2020, cb = TRUE)

# Join the NDI (Powell-Wiley) values to the census tract geometry
DC2020pw <- tract2020DC %>%
  left_join(powell_wiley2020DC$ndi, by = 'GEOID')
DC2020pw <- DC2020pw %>%
  left_join(powell_wiley2020DCi$ndi, by = 'GEOID', suffix = c('_nonimp', '_imp'))
```

```{r powell_wiley_imp_plot, fig.height = 7, fig.width = 7}
# Visualize the NDI (Powell-Wiley) values (2016-2020 5-year ACS) for 
## Washington, D.C., census tracts
## Continuous Index
ggplot() +
  geom_sf(
    data = DC2020pw,
    aes(fill = NDI_nonimp),
    size = 0.2,
    color = 'white'
  ) +
  theme_minimal() +
  scale_fill_viridis_c() +
  labs(fill = 'Index (Continuous)', caption = 'Source: U.S. Census ACS 2016-2020 estimates') +
  ggtitle(
    'Neighborhood Deprivation Index (Powell-Wiley), Non-Imputed',
    subtitle = 'DC census tracts as the referent'
  )

ggplot() +
  geom_sf(
    data = DC2020pw,
    aes(fill = NDI_imp),
    size = 0.2,
    color = 'white'
  ) +
  theme_minimal() +
  scale_fill_viridis_c() +
  labs(fill = 'Index (Continuous)', caption = 'Source: U.S. Census ACS 2016-2020 estimates') +
  ggtitle(
    'Neighborhood Deprivation Index (Powell-Wiley), Imputed',
    subtitle = 'DC census tracts as the referent'
  )

## Categorical Index
### Rename '9-NDI not avail' level as NA for plotting
DC2020pw$NDIQuintNA_nonimp <-
  factor(
    replace(
      as.character(DC2020pw$NDIQuint_nonimp),
      DC2020pw$NDIQuint_nonimp == '9-NDI not avail',
      NA
    ),
    c(levels(DC2020pw$NDIQuint_nonimp)[-6], NA)
  )

ggplot() +
  geom_sf(
    data = DC2020pw,
    aes(fill = NDIQuintNA_nonimp),
    size = 0.2,
    color = 'white'
  ) +
  theme_minimal() +
  scale_fill_viridis_d(guide = guide_legend(reverse = TRUE), na.value = 'grey80') +
  labs(fill = 'Index (Categorical)', caption = 'Source: U.S. Census ACS 2016-2020 estimates') +
  ggtitle(
    'Neighborhood Deprivation Index (Powell-Wiley) Quintiles, Non-Imputed',
    subtitle = 'DC census tracts as the referent'
  )

### Rename '9-NDI not avail' level as NA for plotting
DC2020pw$NDIQuintNA_imp <-
  factor(
    replace(
      as.character(DC2020pw$NDIQuint_imp),
      DC2020pw$NDIQuint_imp == '9-NDI not avail',
      NA
    ),
    c(levels(DC2020pw$NDIQuint_imp)[-6], NA)
  )

ggplot() +
  geom_sf(
    data = DC2020pw,
    aes(fill = NDIQuintNA_imp),
    size = 0.2,
    color = 'white'
  ) +
  theme_minimal() +
  scale_fill_viridis_d(guide = guide_legend(reverse = TRUE), na.value = 'grey80') +
  labs(fill = 'Index (Categorical)', caption = 'Source: U.S. Census ACS 2016-2020 estimates') +
  ggtitle(
    'Neighborhood Deprivation Index (Powell-Wiley) Quintiles, Imputed',
    subtitle = 'DC census tracts as the referent'
  )
```

#### Assign the referent (U.S.-Standardized Metric)

To conduct a contiguous US-standardized index, compute an *NDI* for all states as in the example below that replicates the nationally standardized *NDI* (Powell-Wiley) values (2013-2017 ACS-5) found in [Slotman et al. (2022)](https://doi.org/10.1016/j.dib.2022.108002) and available from a [GIS Portal for Cancer Research](https://gis.cancer.gov/research/files.html#soc-dep) website. To replicate the nationally standardized *NDI* (Powell-Wiley) values (2006-2010 ACS-5) found in [Andrews et al. (2020)](https://doi.org/10.1080/17445647.2020.1750066) change the `year` argument to `2010` (i.e., `year = 2010`).

```{r national_prep, results = 'hide'}
us <- states()
n51 <- c(
  'Commonwealth of the Northern Mariana Islands',
  'Guam',
  'American Samoa',
  'Puerto Rico',
  'United States Virgin Islands'
)
y51 <- us$STUSPS[!(us$NAME %in% n51)]

start_time <- Sys.time() # record start time
powell_wiley2017US <- powell_wiley(state = y51, year = 2017)
end_time <- Sys.time() # record end time
time_srr <- end_time - start_time # Calculate run time
```

```{r national_hist, fig.height = 7, fig.width = 7}
ggplot(powell_wiley2017US$ndi, aes(x = NDI)) +
  geom_histogram(color = 'black', fill = 'white') +
  theme_minimal() +
  ggtitle(
    'Histogram of US-standardized NDI (Powell-Wiley) values (2013-2017)',
    subtitle = 'U.S. census tracts as the referent (including AK, HI, and DC)'
  )
```

The process to compute a US-standardized *NDI* (Powell-Wiley) took about `r round(time_srr, digits = 1)` minutes to run on a machine with the features listed at the end of the vignette.

```{r system}
sessionInfo()
```
