---
title: "ndi: Neighborhood Deprivation Indices"
author: 'Ian D. Buller (GitHub: @idblr)'
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ndi: Neighborhood Deprivation Indices}
  %\VignetteEngine{R.rsp::asis}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = FALSE, fig.show = "hold")
```

Start with the necessary packages for the vignette.

```{r packages, results='hide'}
loadedPackages <- c("dplyr", "ggplot2", "ndi", "tidycensus", "tigris")
invisible(lapply(loadedPackages, library, character.only = TRUE))
options(tigris_use_cache = TRUE)
```

Set your U.S. Census Bureau access key. Follow [this link](http://api.census.gov/data/key_signup.html) to obtain one. Specify your access key in the `messer()` or `powell_wiley()` functions using the `key` argument of the `get_acs()` function from the `tidycensus` package called within each or by using the `census_api_key()` function from the `tidycensus` package before running the `messer()` or `powell_wiley()` functions (see an example of the latter below).

```{r access_key_private, echo=FALSE}
source("../dev/private_key.R")
tidycensus::census_api_key(private_key)
```

```{r access_key_public, eval=FALSE}
tidycensus::census_api_key("...") # INSERT YOUR OWN KEY FROM U.S. CENSUS API
```

### Compute NDI (Messer)

Compute the NDI (Messer) values (2006-2010 5-year ACS) for Georgia, U.S.A., census tracts. This metric is based on [Messer et al. (2006)](https://doi.org/10.1007/s11524-006-9094-x) with the following socio-economic status (SES) variables: 

| Characteristic | SES dimension | ACS table source | Description |
| -------------- | ------------- | ---------------- | ----------- |
| OCC | Occupation | C24030 | Percent males in management, science, and arts occupation |
| CWD | Housing | B25014 | Percent of crowded housing |
| POV | Poverty | B17017 | Percent of households in poverty |
| FHH | Poverty | B25115 | Percent of female headed households with dependents |
| PUB | Poverty | B19058 | Percent of households on public assistance |
| U30 | Poverty | B19001 | Percent households earning <$30,000 per year |
| EDU | Education | B06009 | Percent earning less than a high school education |
| EMP | Employment | B23001 (2010 only); B23025 (2011 onward) | Percent unemployed |

```{r messer}
messer2010GA <- ndi::messer(state = "GA", year = 2010)
```

One output from the `messer()` function is a tibble containing the identification, geographic name, NDI (Messer) values, and raw census characteristics for each tract.

```{r messer_out1}
messer2010GA$ndi
```

A second output from the `messer()` function is the results from the principal component analysis used to compute the NDI (Messer) values.

```{r messer_out2}
messer2010GA$pca
```

A third output from the `messer()` function is a tibble containing a breakdown of the missingness of the census characteristics used to compute the NDI (Messer) values.

```{r messer_out3}
messer2010GA$missing
```

We can visualize the NDI (Messer) values geographically by linking them to spatial information from the `tigris` package and plotting with the `ggplot2` package suite.

```{r messer_plot, fig.height=7, fig.width=7}
# Obtain the 2010 counties from the "tigris" package
county2010GA <- tigris::counties(state = "GA", year = 2010, cb = TRUE)
# Remove first 9 characters from GEOID for compatibility with tigris information
county2010GA$GEOID <- substring(county2010GA$GEO_ID, 10) 

# Obtain the 2010 census tracts from the "tigris" package
tract2010GA <- tigris::tracts(state = "GA", year = 2010, cb = TRUE)
# Remove first 9 characters from GEOID for compatibility with tigris information
tract2010GA$GEOID <- substring(tract2010GA$GEO_ID, 10) 

# Join the NDI (Messer) values to the census tract geometry
GA2010messer <- dplyr::left_join(tract2010GA, messer2010GA$ndi, by = "GEOID")

# Visualize the NDI (Messer) values (2006-2010 5-year ACS) for Georgia, U.S.A., census tracts 
## Continuous Index
ggplot2::ggplot() + 
  ggplot2::geom_sf(data = GA2010messer, 
                   ggplot2::aes(fill = NDI),
                   size = 0.05,
                   color = "white") +
   ggplot2::geom_sf(data = county2010GA,
                   fill = "transparent", 
                   color = "white",
                   size = 0.2) +
  ggplot2::theme_minimal() +
  ggplot2::scale_fill_viridis_c() +
  ggplot2::labs(fill = "Index (Continuous)",
                caption = "Source: U.S. Census ACS 2006-2010 estimates") +
  ggplot2::ggtitle("Neighborhood Deprivation Index (Messer)",
                   subtitle = "GA census tracts as the referent")

## Categorical Index
### Rename "9-NDI not avail" level as NA for plotting
GA2010messer$NDIQuartNA <- factor(replace(as.character(GA2010messer$NDIQuart), 
                                            GA2010messer$NDIQuart == "9-NDI not avail", NA),
                                  c(levels(GA2010messer$NDIQuart)[-5], NA))

ggplot2::ggplot() + 
  ggplot2::geom_sf(data = GA2010messer, 
                   ggplot2::aes(fill = NDIQuartNA),
                   size = 0.05,
                   color = "white") +
   ggplot2::geom_sf(data = county2010GA,
                   fill = "transparent", 
                   color = "white",
                   size = 0.2) +
  ggplot2::theme_minimal() + 
  ggplot2::scale_fill_viridis_d(guide = ggplot2::guide_legend(reverse = TRUE),
                                na.value = "grey80") +
  ggplot2::labs(fill = "Index (Categorical)",
                caption = "Source: U.S. Census ACS 2006-2010 estimates") +
  ggplot2::ggtitle("Neighborhood Deprivation Index (Messer) Quartiles",
                   subtitle = "GA census tracts as the referent")
```

The results above are at the tract level. The NDI (Messer) values can also be calculated at the county level. 

```{r messer_county, fig.height=7, fig.width=7}
messer2010GA_county <- ndi::messer(geo = "county", state = "GA", year = 2010)

# Join the NDI (Messer) values to the county geometry
GA2010messer_county <- dplyr::left_join(county2010GA, messer2010GA_county$ndi, by = "GEOID")

# Visualize the NDI (Messer) values (2006-2010 5-year ACS) for Georgia, U.S.A., counties
## Continuous Index
ggplot2::ggplot() + 
  ggplot2::geom_sf(data = GA2010messer_county, 
                   ggplot2::aes(fill = NDI),
                   size = 0.20,
                   color = "white") +
  ggplot2::theme_minimal() + 
  ggplot2::scale_fill_viridis_c() +
  ggplot2::labs(fill = "Index (Continuous)",
                caption = "Source: U.S. Census ACS 2006-2010 estimates") +
  ggplot2::ggtitle("Neighborhood Deprivation Index (Messer)",
                   subtitle = "GA counties as the referent")

## Categorical Index

### Rename "9-NDI not avail" level as NA for plotting
GA2010messer_county$NDIQuartNA <- factor(replace(as.character(GA2010messer_county$NDIQuart), 
                                            GA2010messer_county$NDIQuart == "9-NDI not avail", NA),
                                         c(levels(GA2010messer_county$NDIQuart)[-5], NA))

ggplot2::ggplot() + 
  ggplot2::geom_sf(data = GA2010messer_county, 
                   ggplot2::aes(fill = NDIQuartNA),
                   size = 0.20,
                   color = "white") +
  ggplot2::theme_minimal() + 
  ggplot2::scale_fill_viridis_d(guide = ggplot2::guide_legend(reverse = TRUE),
                                na.value = "grey80") +
  ggplot2::labs(fill = "Index (Categorical)",
                caption = "Source: U.S. Census ACS 2006-2010 estimates") +
  ggplot2::ggtitle("Neighborhood Deprivation Index (Messer) Quartiles",
                   subtitle = "GA counties as the referent")
```

### Compute NDI (Powell-Wiley)

Compute the NDI (Powell-Wiley) values (2016-2020 5-year ACS) for Maryland, Virginia, Washington, D.C., and West Virginia, U.S.A., census tracts. This metric is based on [Andrews et al. (2020)](https://doi.org/10.1080/17445647.2020.1750066) and [Slotman et al. (2022)](https://doi.org/10.1016/j.dib.2022.108002) with socio-economic status (SES) variables chosen by [Roux and Mair (2010)](https://doi.org/10.1111/j.1749-6632.2009.05333.x):

| Characteristic | SES dimension | ACS table source | Description |
| -------------- | ------------- | ---------------- | ----------- |
| MedHHInc | Wealth and income | B19013 | Median household income (dollars) |
| PctRecvIDR | Wealth and income | B19054 | Percent of households receiving dividends, interest, or rental income |
| PctPubAsst | Wealth and income | B19058 | Percent of households receiving public assistance |
| MedHomeVal | Wealth and income | B25077 | Median home value (dollars) |
| PctMgmtBusSciArt | Occupation | C24060 | Percent in a management, business, science, or arts occupation |
| PctFemHeadKids | Occupation | B11005 | Percent of households that are female headed with any children under 18 years |
| PctOwnerOcc | Housing conditions | DP04 | Percent of housing units that are owner occupied |
| PctNoPhone | Housing conditions | DP04 | Percent of households without a telephone |
| PctNComPlmb | Housing conditions | DP04 | Percent of households without complete plumbing facilities |
| PctEducHSPlus | Education | S1501 | Percent with a high school degree or higher (population 25 years and over) |
| PctEducBchPlus | Education | S1501 | Percent with a college degree or higher (population 25 years and over) |
| PctFamBelowPov | Wealth and income | S1702 | Percent of families with incomes below the poverty level |
| PctUnempl | Occupation | S2301 | Percent unemployed |

More information about the [codebook](https://gis.cancer.gov/research/NeighDeprvIndex_Methods.pdf) and [computation](https://gis.cancer.gov/research/NeighDeprvIndex_Methods.pdf) of the NDI (Powell-Wiley) can be found on a [GIS Portal for Cancer Research](https://gis.cancer.gov/research/files.html#soc-dep) website.

```{r powell_wiley}
powell_wiley2020DMVW <- ndi::powell_wiley(state = c("DC", "MD", "VA", "WV"), year = 2020)
```

One output from the `powell_wiley()` function is a tibble containing the identification, geographic name, NDI (Powell-Wiley) values, and raw census characteristics for each tract.

```{r powell_wiley_out1}
powell_wiley2020DMVW$ndi
```

A second output from the `powell_wiley()` function is the results from the principal component analysis used to compute the NDI (Powell-Wiley) values.

```{r powell_wiley_out2}
powell_wiley2020DMVW$pca
```

A third output from the `powell_wiley()` function is a tibble containing a breakdown of the missingness of the census characteristics used to compute the NDI (Powell-Wiley) values.

```{r powell_wiley_out3}
powell_wiley2020DMVW$missing
```

A fourth output from the `powell_wiley()` function is a character string or numeric value of a standardized Cronbach's alpha. A value greater than 0.7 is desired.

```{r powell_wiley_out4}
powell_wiley2020DMVW$cronbach
```

We can visualize the NDI (Powell-Wiley) values geographically by linking them to spatial information from the `tigris` package and plotting with the `ggplot2` package suite.

```{r powell_wiley_plot, fig.height=4, fig.width=7}
# Obtain the 2020 census tracts from the "tigris" package
state2020 <- tigris::states(cb = TRUE)
state2020DMVW <- state2020[state2020$STUSPS %in% c("DC", "MD", "VA", "WV"), ]
tract2020D <- tigris::tracts(state = "DC", year = 2020, cb = TRUE)
tract2020M <- tigris::tracts(state = "MD", year = 2020, cb = TRUE)
tract2020V <- tigris::tracts(state = "VA", year = 2020, cb = TRUE)
tract2020W <- tigris::tracts(state = "WV", year = 2020, cb = TRUE)
tracts2020DMVW <- rbind(tract2020D, tract2020M, tract2020V, tract2020W)

# Join the NDI (Powell-Wiley) values to the census tract geometry
DMVW2020pw <- dplyr::left_join(tracts2020DMVW, powell_wiley2020DMVW$ndi, by = "GEOID")

# Visualize the NDI (Powell-Wiley) values (2016-2020 5-year ACS) 
## Maryland, Virginia, Washington, D.C., and West Virginia, U.S.A., census tracts 
## Continuous Index
ggplot2::ggplot() + 
  ggplot2::geom_sf(data = DMVW2020pw, 
                   ggplot2::aes(fill = NDI), 
                   color = NA) +
  ggplot2::geom_sf(data = state2020DMVW,
                   fill = "transparent", 
                   color = "white") +
  ggplot2::theme_minimal() + 
  ggplot2::scale_fill_viridis_c(na.value = "grey80") +
  ggplot2::labs(fill = "Index (Continuous)",
                caption = "Source: U.S. Census ACS 2016-2020 estimates")+
  ggplot2::ggtitle("Neighborhood Deprivation Index (Powell-Wiley)",
                   subtitle = "DC, MD, VA, and WV tracts as the referent")

## Categorical Index (Population-weighted quintiles)
### Rename "9-NDI not avail" level as NA for plotting
DMVW2020pw$NDIQuintNA <- factor(replace(as.character(DMVW2020pw$NDIQuint), 
                                        DMVW2020pw$NDIQuint == "9-NDI not avail", NA),
                                c(levels(DMVW2020pw$NDIQuint)[-6], NA))

ggplot2::ggplot() + 
  ggplot2::geom_sf(data = DMVW2020pw, 
                   ggplot2::aes(fill = NDIQuintNA), 
                   color = NA) +
  ggplot2::geom_sf(data = state2020DMVW,
                   fill = "transparent", 
                   color = "white") +
  ggplot2::theme_minimal() + 
  ggplot2::scale_fill_viridis_d(guide = ggplot2::guide_legend(reverse = TRUE),
                                na.value = "grey80") +
  ggplot2::labs(fill = "Index (Categorical)",
                caption = "Source: U.S. Census ACS 2016-2020 estimates")+
  ggplot2::ggtitle("Neighborhood Deprivation Index (Powell-Wiley) Population-weighted Quintiles",
                   subtitle = "DC, MD, VA, and WV tracts as the referent")
```

Like the NDI (Messer), we also compute county-level NDI (Powell-Wiley).

```{r powell_wiley_county, fig.height=4, fig.width=7}
# Obtain the 2020 counties from the "tigris" package
county2020DMVW <- tigris::counties(state = c("DC", "MD", "VA", "WV"), year = 2020, cb = TRUE)

# NDI (Powell-Wiley) at the county level (2016-2020)
powell_wiley2020DMVW_county <- ndi::powell_wiley(geo = "county",
                                                 state = c("DC", "MD", "VA", "WV"),
                                                 year = 2020)

# Join the NDI (Powell-Wiley) values to the county geometry
DMVW2020pw_county <- dplyr::left_join(county2020DMVW, powell_wiley2020DMVW_county$ndi, by = "GEOID")

# Visualize the NDI (Powell-Wiley) values (2016-2020 5-year ACS)
## Maryland, Virginia, Washington, D.C., and West Virginia, U.S.A., counties
## Continuous Index
ggplot2::ggplot() + 
  ggplot2::geom_sf(data = DMVW2020pw_county, 
                   ggplot2::aes(fill = NDI),
                   size = 0.20,
                   color = "white") +
  ggplot2::theme_minimal() + 
  ggplot2::scale_fill_viridis_c() +
  ggplot2::labs(fill = "Index (Continuous)",
                caption = "Source: U.S. Census ACS 2016-2020 estimates") +
  ggplot2::ggtitle("Neighborhood Deprivation Index (Powell-Wiley)",
                   subtitle = "DC, MD, VA, and WV counties as the referent")

## Categorical Index

### Rename "9-NDI not avail" level as NA for plotting
DMVW2020pw_county$NDIQuintNA <- factor(replace(as.character(DMVW2020pw_county$NDIQuint), 
                                            DMVW2020pw_county$NDIQuint == "9-NDI not avail", NA),
                                         c(levels(DMVW2020pw_county$NDIQuint)[-6], NA))

ggplot2::ggplot() + 
  ggplot2::geom_sf(data = DMVW2020pw_county, 
                   ggplot2::aes(fill = NDIQuint),
                   size = 0.20,
                   color = "white") +
  ggplot2::theme_minimal() + 
  ggplot2::scale_fill_viridis_d(guide = ggplot2::guide_legend(reverse = TRUE),
                                na.value = "grey80") +
  ggplot2::labs(fill = "Index (Categorical)",
                caption = "Source: U.S. Census ACS 2016-2020 estimates") +
  ggplot2::ggtitle("Neighborhood Deprivation Index (Powell-Wiley) Population-weighted Quintiles",
                   subtitle = "DC, MD, VA, and WV counties as the referent")
```

### Advanced Features

#### Imputing missing census variables

In the `messer()` and `powell_wiley()` functions, missing census characteristics can be imputed using the `missing` and `impute` arguments of the `pca()` function in the `psych` package called within the `messer()` and `powell_wiley()` functions. Impute values using the logical `imp` argument (currently only calls `impute = "median"` by default, which assigns the median values of each missing census variable for a geography).

```{r powell_wiley_imp, results='hide', fig.height=7, fig.width=7}
powell_wiley2020DC <- ndi::powell_wiley(state = "DC", year = 2020) # without imputation
powell_wiley2020DCi <- ndi::powell_wiley(state = "DC", year = 2020, imp = TRUE) # with imputation

table(is.na(powell_wiley2020DC$ndi$NDI)) # n=13 tracts without NDI (Powell-Wiley) values
table(is.na(powell_wiley2020DCi$ndi$NDI)) # n=0 tracts without NDI (Powell-Wiley) values

# Obtain the 2020 census tracts from the "tigris" package
tract2020DC <- tigris::tracts(state = "DC", year = 2020, cb = TRUE)

# Join the NDI (Powell-Wiley) values to the census tract geometry
DC2020pw <- dplyr::left_join(tract2020DC, powell_wiley2020DC$ndi, by = "GEOID")
DC2020pw <- dplyr::left_join(DC2020pw, powell_wiley2020DCi$ndi, by = "GEOID", suffix = c("_nonimp", "_imp"))

# Visualize the NDI (Powell-Wiley) values (2016-2020 5-year ACS) for Washington, D.C., census tracts
## Continuous Index
ggplot2::ggplot() + 
  ggplot2::geom_sf(data = DC2020pw, 
                   ggplot2::aes(fill = NDI_nonimp),
                   size = 0.2,
                   color = "white") +
  ggplot2::theme_minimal() + 
  ggplot2::scale_fill_viridis_c() +
  ggplot2::labs(fill = "Index (Continuous)",
                caption = "Source: U.S. Census ACS 2016-2020 estimates") +
  ggplot2::ggtitle("Neighborhood Deprivation Index (Powell-Wiley), Non-Imputed",
                   subtitle = "DC census tracts as the referent")

ggplot2::ggplot() + 
  ggplot2::geom_sf(data = DC2020pw, 
                   ggplot2::aes(fill = NDI_imp),
                   size = 0.2,
                   color = "white") +
  ggplot2::theme_minimal() + 
  ggplot2::scale_fill_viridis_c() +
  ggplot2::labs(fill = "Index (Continuous)",
                caption = "Source: U.S. Census ACS 2016-2020 estimates") +
  ggplot2::ggtitle("Neighborhood Deprivation Index (Powell-Wiley), Imputed",
                   subtitle = "DC census tracts as the referent")

## Categorical Index
### Rename "9-NDI not avail" level as NA for plotting
DC2020pw$NDIQuintNA_nonimp <- factor(replace(as.character(DC2020pw$NDIQuint_nonimp), 
                                            DC2020pw$NDIQuint_nonimp == "9-NDI not avail", NA),
                                         c(levels(DC2020pw$NDIQuint_nonimp)[-6], NA))

ggplot2::ggplot() + 
  ggplot2::geom_sf(data = DC2020pw, 
                   ggplot2::aes(fill = NDIQuintNA_nonimp),
                   size = 0.2,
                   color = "white") +
  ggplot2::theme_minimal() + 
  ggplot2::scale_fill_viridis_d(guide = ggplot2::guide_legend(reverse = TRUE),
                                na.value = "grey80") +
  ggplot2::labs(fill = "Index (Categorical)",
                caption = "Source: U.S. Census ACS 2016-2020 estimates") +
  ggplot2::ggtitle("Neighborhood Deprivation Index (Powell-Wiley) Quintiles, Non-Imputed",
                   subtitle = "DC census tracts as the referent")

### Rename "9-NDI not avail" level as NA for plotting
DC2020pw$NDIQuintNA_imp <- factor(replace(as.character(DC2020pw$NDIQuint_imp), 
                                            DC2020pw$NDIQuint_imp == "9-NDI not avail", NA),
                                      c(levels(DC2020pw$NDIQuint_imp)[-6], NA))

ggplot2::ggplot() + 
  ggplot2::geom_sf(data = DC2020pw, 
                   ggplot2::aes(fill = NDIQuintNA_imp),
                   size = 0.2,
                   color = "white") +
  ggplot2::theme_minimal() + 
  ggplot2::scale_fill_viridis_d(guide = ggplot2::guide_legend(reverse = TRUE),
                                na.value = "grey80") +
  ggplot2::labs(fill = "Index (Categorical)",
                caption = "Source: U.S. Census ACS 2016-2020 estimates") +
  ggplot2::ggtitle("Neighborhood Deprivation Index (Powell-Wiley) Quintiles, Imputed",
                   subtitle = "DC census tracts as the referent")
```

#### Assign the referent (U.S.-Standardized Metric)

To conduct a contiguous US-standardized index, compute an NDI for all states as in the example below that replicates the nationally standardized NDI (Powell-Wiley) values (2013-2017 ACS-5) found in [Slotman et al. (2022)](https://doi.org/10.1016/j.dib.2022.108002) and available from a [GIS Portal for Cancer Research](https://gis.cancer.gov/research/files.html#soc-dep) website. To replicate the nationally standardized NDI (Powell-Wiley) values (2006-2010 ACS-5) found in [Andrews et al. (2020)](https://doi.org/10.1080/17445647.2020.1750066) change the `year` argument to `2010` (i.e., `year = 2010`).

```{r national}
us <- tigris::states()
n51 <- c("Commonwealth of the Northern Mariana Islands", "Guam", "American Samoa",
         "Puerto Rico", "United States Virgin Islands")
y51 <- us$STUSPS[!(us$NAME %in% n51)]

start_time <- Sys.time() # record start time
powell_wiley2017US <- ndi::powell_wiley(state = y51, year = 2017)
end_time <- Sys.time() # record end time
time_srr <- end_time - start_time # Calculate run time
powell_wiley2017US
```

```{r national_hist, fig.height=7, fig.width=7}
ggplot2::ggplot(powell_wiley2017US$ndi, 
                ggplot2::aes(x = NDI)) +
  ggplot2::geom_histogram(color = "black",
                          fill = "white") + 
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("Histogram of US-standardized NDI (Powell-Wiley) values (2013-2017)",
                   subtitle = "U.S. census tracts as the referent (including AK, HI, and DC)")
```

The process to compute a US-standardized NDI (Powell-Wiley) took about `r round(time_srr, digits = 1)` minutes to run on a machine with the features listed at the end of the vignette.

### Additional metrics socio-economic deprivation and disparity 

Since version v0.1.1, the `ndi` package can compute additional metrics of socio-economic deprivation and disparity beyond neighborhood deprivation indices, including:

1. `anthopolos()` function that computes the Racial Isolation Index (RI) based on [Anthopolos et al. (2011)](https://www.doi.org/10.1016/j.sste.2011.06.002) with data from the ACS-5.
2. `bravo()` function that computes the Educational Isolation Index (EI) based on [Bravo et al. (2021)](https://www.doi.org/10.3390/ijerph18179384) with data from the ACS-5.
3. `gini()` function that retrieves the Gini Index based on [Gini (1921)](https://www.doi.org/10.2307/2223319) from the ACS-5. 
4. `krieger()` function that computes the Index of Concentration at the Extremes based on based on [Feldman et al. (2015)](https://www.doi.org/10.1136/jech-2015-205728) and [Krieger et al. (2016)](https://www.doi.org/10.2105/AJPH.2015.302955) with data from the ACS-5. 

#### Compute Racial Isolation Index (RI)

Compute the RI (Anthopolos) values (2006-2010 5-year ACS) for North Carolina, U.S.A., census tracts. This metric is based on [Anthopolos et al. (2011)](https://www.doi.org/10.1016/j.sste.2011.06.002) that assessed the racial isolation of the population that identifies as non-Hispanic or Latino, Black or African American alone. Multiple racial/ethnic subgroups are available in the `anthopolos()` function, including:

| ACS table source | racial/ethnic subgroup | character for `subgroup` argument |
| -------------- | ------------- | ---------------- |
| B03002_002 | not Hispanic or Latino | NHoL |
| B03002_003 | not Hispanic or Latino, white alone | NHoLW |
| B03002_004 | not Hispanic or Latino, Black or African American alone | NHoLB |
| B03002_005 | not Hispanic or Latino, American Indian and Alaska Native alone | NHoLAIAN |
| B03002_006 | not Hispanic or Latino, Asian alone | NHoLA |
| B03002_007 | not Hispanic or Latino, Native Hawaiian and Other Pacific Islander alone | NHoLNHOPI |
| B03002_008 | not Hispanic or Latino, some other race alone | NHoLSOR |
| B03002_009 | not Hispanic or Latino, two or more races | NHoLTOMR |
| B03002_010 | not Hispanic or Latino, two races including some other race | NHoLTRiSOR |
| B03002_011 | not Hispanic or Latino, two races excluding some other race, and three or more races | NHoLTReSOR |
| B03002_012 | nispanic or Latino | HoL |
| B03002_013 | Hispanic or Latino, white alone | HoLW |
| B03002_014 | Hispanic or Latino, Black or African American alone | HoLB |
| B03002_015 | Hispanic or Latino, American Indian and Alaska Native alone | HoLAIAN |
| B03002_016 | Hispanic or Latino, Asian alone | HoLA |
| B03002_017 | Hispanic or Latino, Native Hawaiian and other Pacific Islander alone | HoLNHOPI |
| B03002_018 | Hispanic or Latino, some other race alone | HoLSOR |
| B03002_019 | Hispanic or Latino, two or more races | HoLTOMR |
| B03002_020 | Hispanic or Latino, two races including some other race | HoLTRiSOR |
| B03002_021 | Hispanic or Latino, two races excluding some other race, and three or more races | HoLTReSOR |

A census geography (and its neighbors) that has nearly all of its population who identify with the specified race/ethnicity subgroup(s) (e.g., Not Hispanic or Latino, Black or African American alone) will have an RI value close to 1. In contrast, a census geography (and its neighbors) that is nearly none of its population who identify with the specified race/ethnicity subgroup(s) (e.g., not Not Hispanic or Latino, Black or African American alone) will have an RI value close to 0.

```{r anthopolos, fig.height=4, fig.width=7}
anthopolos2010NC <- ndi::anthopolos(state = "NC", year = 2010, subgroup = "NHoLB")

# Obtain the 2010 census tracts from the "tigris" package
tract2010NC <- tigris::tracts(state = "NC", year = 2010, cb = TRUE)
# Remove first 9 characters from GEOID for compatibility with tigris information
tract2010NC$GEOID <- substring(tract2010NC$GEO_ID, 10) 

# Obtain the 2010 counties from the "tigris" package
county2010NC <- tigris::counties(state = "NC", year = 2010, cb = TRUE)

# Join the RI (Anthopolos) values to the census tract geometry
NC2010anthopolos <- dplyr::left_join(tract2010NC, anthopolos2010NC$ri, by = "GEOID")

# Visualize the RI (Anthopolos) values (2006-2010 5-year ACS) for North Carolina, U.S.A., census tracts 
ggplot2::ggplot() + 
  ggplot2::geom_sf(data = NC2010anthopolos, 
                   ggplot2::aes(fill = RI),
                   size = 0.05,
                   color = "white") +
   ggplot2::geom_sf(data = county2010NC,
                   fill = "transparent", 
                   color = "white",
                   size = 0.2) +
  ggplot2::theme_minimal() +
  ggplot2::scale_fill_viridis_c() +
  ggplot2::labs(fill = "Index (Continuous)",
                caption = "Source: U.S. Census ACS 2006-2010 estimates") +
  ggplot2::ggtitle("Racial Isolation Index (Anthopolos), non-Hispanic Black",
                   subtitle = "NC census tracts (not corrected for edge effects)")
```

The current version of the `ndi` package does not correct for edge effects (e.g., census geographies along the specified spatial extent border, coastline, or U.S.-Mexico / U.S.-Canada border) may have few neighboring census geographies, and RI values in these census geographies may be unstable. A stop-gap solution for the former source of edge effect is to compute the RI for neighboring census geographies (i.e., the states bordering a study area of interest) and then use the estimates of the study area of interest.

```{r anthopolos_edge, fig.height=4, fig.width=7}
# Compute RI for all census tracts in neighboring states
anthopolos2010GNSTV <- ndi::anthopolos(state = c("GA", "NC", "SC", "TN", "VA"),
                                     year = 2010, subgroup = "NHoLB")

# Crop to only North Carolina, U.S.A. census tracts
anthopolos2010NCe <- anthopolos2010GNSTV$ri[anthopolos2010GNSTV$ri$GEOID %in% anthopolos2010NC$ri$GEOID, ]

# Obtain the 2010 census tracts from the "tigris" package
tract2010NC <- tigris::tracts(state = "NC", year = 2010, cb = TRUE)
# Remove first 9 characters from GEOID for compatibility with tigris information
tract2010NC$GEOID <- substring(tract2010NC$GEO_ID, 10) 

# Obtain the 2010 counties from the "tigris" package
county2010NC <- tigris::counties(state = "NC", year = 2010, cb = TRUE)

# Join the RI (Anthopolos) values to the census tract geometry
edgeNC2010anthopolos <- dplyr::left_join(tract2010NC, anthopolos2010NCe, by = "GEOID")

# Visualize the RI (Anthopolos) values (2006-2010 5-year ACS) for North Carolina, U.S.A., census tracts 
ggplot2::ggplot() + 
  ggplot2::geom_sf(data = edgeNC2010anthopolos, 
                   ggplot2::aes(fill = RI),
                   size = 0.05,
                   color = "white") +
   ggplot2::geom_sf(data = county2010NC,
                   fill = "transparent", 
                   color = "white",
                   size = 0.2) +
  ggplot2::theme_minimal() +
  ggplot2::scale_fill_viridis_c() +
  ggplot2::labs(fill = "Index (Continuous)",
                caption = "Source: U.S. Census ACS 2006-2010 estimates") +
  ggplot2::ggtitle("Racial Isolation Index (Anthopolos), non-Hispanic Black",
                   subtitle = "NC census tracts (corrected for interstate edge effects)")
```

#### Compute Educational Isolation Index (EI)

Compute the EI (Bravo) values (2006-2010 5-year ACS) for North Carolina, U.S.A., census tracts. This metric is based on [Bravo et al. (2021)](https://www.doi.org/10.3390/ijerph18179384) that assessed the educational isolation of the population without a four-year college degree. Multiple educational attainment categories are available in the `bravo()` function, including:

| ACS table source | educational attainment category | character for `subgroup` argument |
| -------------- | ------------- | ---------------- |
| B06009_002 | less than high school graduate | LtHS |
| B06009_003 | high school graduate (includes equivalency) | HSGiE |
| B06009_004 | some college or associate's degree | SCoAD |
| B06009_005 | Bachelor's degree | BD |
| B06009_006 | graduate or professional degree | GoPD |

A census geography (and its neighbors) that has nearly all of its population with the specified educational attainment category (e.g., a four-year college degree or more) will have an EI value close to 1. In contrast, a census geography (and its neighbors) that is nearly none of its population with the specified educational attainment category (e.g., with a four-year college degree) will have an EI value close to 0.

```{r bravo, fig.height=4, fig.width=7}
bravo2010NC <- ndi::bravo(state = "NC", year = 2010, subgroup = c("LtHS", "HSGiE", "SCoAD"))

# Obtain the 2010 census tracts from the "tigris" package
tract2010NC <- tigris::tracts(state = "NC", year = 2010, cb = TRUE)
# Remove first 9 characters from GEOID for compatibility with tigris information
tract2010NC$GEOID <- substring(tract2010NC$GEO_ID, 10) 

# Obtain the 2010 counties from the "tigris" package
county2010NC <- tigris::counties(state = "NC", year = 2010, cb = TRUE)

# Join the RI (Bravo) values to the census tract geometry
NC2010bravo <- dplyr::left_join(tract2010NC, bravo2010NC$ei, by = "GEOID")

# Visualize the RI (Bravo) values (2006-2010 5-year ACS) for North Carolina, U.S.A., census tracts 
ggplot2::ggplot() + 
  ggplot2::geom_sf(data = NC2010bravo, 
                   ggplot2::aes(fill = EI),
                   size = 0.05,
                   color = "white") +
   ggplot2::geom_sf(data = county2010NC,
                   fill = "transparent", 
                   color = "white",
                   size = 0.2) +
  ggplot2::theme_minimal() +
  ggplot2::scale_fill_viridis_c() +
  ggplot2::labs(fill = "Index (Continuous)",
                caption = "Source: U.S. Census ACS 2006-2010 estimates") +
  ggplot2::ggtitle("Educational Isolation Index (Bravo), without a four-year college degree",
                   subtitle = "NC census tracts (not corrected for edge effects)")
```

Can correct one source of edge effect in the same manner as shown for the RI (Anthopolos) metric.

#### Retrieve the Gini Index

Retrieve the Gini Index values (2006-2010 5-year ACS) for North Carolina, U.S.A., census tracts. This metric is based on [Gini (1921)](https://www.doi.org/10.2307/2223319), and the `gini()` function retrieves the estimate from the ACS-5. 

According to the [U.S. Census Bureau](https://www.census.gov/topics/income-poverty/income-inequality/about/metrics/gini-index.html): "The Gini Index is a summary measure of income inequality. The Gini coefficient incorporates the detailed shares data into a single statistic, which summarizes the dispersion of income across the entire income distribution. The Gini coefficient ranges from 0, indicating perfect equality (where everyone receives an equal share), to 1, perfect inequality (where only one recipient or group of recipients receives all the income). The Gini is based on the difference between the Lorenz curve (the observed cumulative income distribution) and the notion of a perfectly equal income distribution."

```{r gini, fig.height=4, fig.width=7}
gini2010NC <- ndi::gini(state = "NC", year = 2010)

# Obtain the 2010 census tracts from the "tigris" package
tract2010NC <- tigris::tracts(state = "NC", year = 2010, cb = TRUE)
# Remove first 9 characters from GEOID for compatibility with tigris information
tract2010NC$GEOID <- substring(tract2010NC$GEO_ID, 10) 

# Obtain the 2010 counties from the "tigris" package
county2010NC <- tigris::counties(state = "NC", year = 2010, cb = TRUE)

# Join the Gini Index values to the census tract geometry
NC2010gini <- dplyr::left_join(tract2010NC, gini2010NC$gini, by = "GEOID")

# Visualize the Gini Index values (2006-2010 5-year ACS) for North Carolina, U.S.A., census tracts 
ggplot2::ggplot() + 
  ggplot2::geom_sf(data = NC2010gini, 
                   ggplot2::aes(fill = gini),
                   size = 0.05,
                   color = "white") +
   ggplot2::geom_sf(data = county2010NC,
                   fill = "transparent", 
                   color = "white",
                   size = 0.2) +
  ggplot2::theme_minimal() +
  ggplot2::scale_fill_viridis_c() +
  ggplot2::labs(fill = "Index (Continuous)",
                caption = "Source: U.S. Census ACS 2006-2010 estimates") +
  ggplot2::ggtitle("Gini Index",
                   subtitle = "NC census tracts")
```

### Index of Concentration at the Extremes (ICE)

Compute the Index of Concentration at the Extremes values (2006-2010 5-year ACS) for Wayne County, Michigan, U.S.A., census tracts. Wayne County is the home of Detroit, Michigan, a highly segregated city in the U.S. This metric is based on [Feldman et al. (2015)](https://www.doi.org/10.1136/jech-2015-205728) and [Krieger et al. (2016)](https://www.doi.org/10.2105/AJPH.2015.302955) who expanded the metric designed by Massey in a chapter of [Booth & Crouter (2001)](https://www.doi.org/10.4324/9781410600141) initially designed for residential segregation. The `krieger()` function computes five ICE metrics using the following ACS-5 groups:

| ACS table group | ICE metric | Comparison
| -------------- | ------------- | ---------------- |
| B19001 | Income, "ICE_inc"| 80th income percentile vs. 20th income percentile |
| B15002 | Education, "ICE_edu"| less than high school vs. four-year college degree or more |
| B03002 | Race/Ethnicity, "ICE_rewb"| 80th income percentile vs. 20th income percentile |
| B19001 & B19001B & B19001H | Income and race/ethnicity combined, "ICE_wbinc" | white non-Hispanic in 80th income percentile vs. black alone (including Hispanic) in 20th income percentile |
| B19001 & B19001H | Income and race/ethnicity combined, "ICE_wpcinc"| white non-Hispanic in 80th income percentile vs. white non-Hispanic in 20th income percentile |

ICE metrics can range in value from −1 (most deprived) to 1 (most privileged). A value of 0 can thus represent two possibilities: (1) none of the residents are in the most privileged or most deprived categories, or (2) an equal number of persons are in the most privileged and most deprived categories, and in both cases indicates that the area is not dominated by extreme concentrations of either of the two groups.

```{r krieger, fig.height=5.5, fig.width=7}
ice2020WC <- krieger(state = "MI", county = "Wayne", year = 2010)

# Obtain the 2010 census tracts from the "tigris" package
tract2010WC <- tigris::tracts(state = "MI", county = "Wayne", year = 2010, cb = TRUE)
# Remove first 9 characters from GEOID for compatibility with tigris information
tract2010WC$GEOID <- substring(tract2010WC$GEO_ID, 10) 

# Join the ICEs (Krieger) values to the census tract geometry
ice2020WC <- dplyr::left_join(tract2010WC, ice2020WC$ice, by = "GEOID")

# Plot ICE for Income
ggplot2::ggplot() + 
  ggplot2::geom_sf(data = ice2020WC, 
                   ggplot2::aes(fill = ICE_inc),
                   color = "white",
                   size = 0.05) +
  ggplot2::theme_bw() + 
  ggplot2::scale_fill_gradient2(low = "#998ec3", mid = "#f7f7f7", high = "#f1a340", limits = c(-1,1)) +
  ggplot2::labs(fill = "Index (Continuous)",
                caption = "Source: U.S. Census ACS 2006-2010 estimates")+
  ggplot2::ggtitle("Index of Concentration at the Extremes\nIncome (Krieger)",
                   subtitle = "80th income percentile vs. 20th income percentile")

# Plot ICE for Education
ggplot2::ggplot() + 
  ggplot2::geom_sf(data = ice2020WC, 
                   ggplot2::aes(fill = ICE_edu),
                   color = "white",
                   size = 0.05) +
  ggplot2::theme_bw() + 
  ggplot2::scale_fill_gradient2(low = "#998ec3", mid = "#f7f7f7", high = "#f1a340", limits = c(-1,1)) +
  ggplot2::labs(fill = "Index (Continuous)",
                caption = "Source: U.S. Census ACS 2006-2010 estimates")+
  ggplot2::ggtitle("Index of Concentration at the Extremes\nEducation (Krieger)",
                   subtitle = "less than high school vs. four-year college degree or more")

# Plot ICE for Race/Ethnicity
ggplot2::ggplot() + 
  ggplot2::geom_sf(data = ice2020WC, 
                   ggplot2::aes(fill = ICE_rewb),
                   color = "white",
                   size = 0.05) +
  ggplot2::theme_bw() + 
  ggplot2::scale_fill_gradient2(low = "#998ec3", mid = "#f7f7f7", high = "#f1a340", limits = c(-1, 1)) +
  ggplot2::labs(fill = "Index (Continuous)",
                caption = "Source: U.S. Census ACS 2006-2010 estimates")+
  ggplot2::ggtitle("Index of Concentration at the Extremes\nRace/Ethnicity (Krieger)",
                   subtitle = "white non-Hispanic vs. black non-Hispanic")

# Plot ICE for Income and Race/Ethnicity Combined
## white non-Hispanic in 80th income percentile vs. black (including Hispanic) in 20th income percentile
ggplot2::ggplot() + 
  ggplot2::geom_sf(data = ice2020WC, 
                   ggplot2::aes(fill = ICE_wbinc),
                   color = "white",
                   size = 0.05) +
  ggplot2::theme_bw() + 
  ggplot2::scale_fill_gradient2(low = "#998ec3", mid = "#f7f7f7", high = "#f1a340", limits = c(-1, 1)) +
  ggplot2::labs(fill = "Index (Continuous)",
                caption = "Source: U.S. Census ACS 2006-2010 estimates")+
  ggplot2::ggtitle("Index of Concentration at the Extremes\nIncome & race/ethnicity combined (Krieger)",
                   subtitle = "white non-Hispanic in 80th inc ptcl vs. black alone in 20th inc pctl")

# Plot ICE for Income and Race/Ethnicity Combined
## white non-Hispanic in 80th income percentile vs. white non-Hispanic in 20th income percentile
ggplot2::ggplot() + 
  ggplot2::geom_sf(data = ice2020WC, 
                   ggplot2::aes(fill = ICE_wpcinc),
                   color = "white",
                   size = 0.05) +
  ggplot2::theme_bw() + 
  ggplot2::scale_fill_gradient2(low = "#998ec3", mid = "#f7f7f7", high = "#f1a340", limits = c(-1, 1)) +
  ggplot2::labs(fill = "Index (Continuous)",
                caption = "Source: U.S. Census ACS 2006-2010 estimates")+
  ggplot2::ggtitle("Index of Concentration at the Extremes\nIncome & race/ethnicity combined (Krieger)",
                   subtitle = "white non-Hispanic (WNH) in 80th inc pctl vs. WNH in 20th inc pctl")
```

```{r system}
sessionInfo()
```
